Runtime-Assisted Convergence in Replicated Data Types

### 2021-11-20 evaluation.tex
Edits

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{rope}), we decided to adopt the simplest representation of
a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. 
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.  \quark's \C{write} is a function of type
$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
\C{doc} in the current application.

@@

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. We conduct the experiments on a
three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 60 and measure the impact of the increased
concurrency on write latency. Fig.~\ref{fig:latency} shows the median
latency values. As evident from the figure, latency of \quark writes
constant with the median latency around 0.008s. The maximum latency
ever measured for a \quark writes is 0.016s. Notably, the latency
values are an order of magnitude less than the time between
consecutive edits (0.25s), which lets users see their edits almost
immediately. Fig.~\ref{fig:latency} also plots the latency values for
the baseline ``SC'' approach which achieves convergence by
synchronizing each operation, i.e., executing under strong
consistency. The SC write latency increases super-linearly with the
median value of 1.76s for 60 concurrent editors. The maximum latency
measured for an SC write is 9.02s, which is considerably more than the
inter-edit latency of 0.25s. The experiment confirms the theory that
convergence does not come at the cost of latency in \quark.

@@

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with progression of time,
$B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't stop it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment and analyzed to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators in the multiples of 3 until 30. While
increased staleness is inevitable in our approach due to synchronized
merges, we believe the increase can be contained by choosing the order
of merges to avoid the ``starvation'' of some branches. Note that
relaxing the linearizability constraint on merges would completely
eliminate the staleness overhead, but the resultant Git-like system
fails to converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits anamolous version history graphs where two branches can
  have the same set of commits and yet differ in their final version.
  Supplementary material describes two such cases we observed on
  Github.}.

>>>>

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{boehm95}), we decided to adopt the simplest representation
of a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. 
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.
Thus, \quark's \C{write} is a function of type $\texttt{Value}
\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
current application.

@@

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. We conduct the experiments on a
three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 60 and measure the impact of the increased
concurrency on write latency. Fig.~\ref{fig:latency} shows the median
latency values. As evident from the figure, latency of \quark writes
remain more-or-less constant with the median latency around 0.008s.
The maximum latency ever measured for a \quark writes is 0.016s.
Notably, latency remains an order of magnitude less than the time
between consecutive edits (0.25s), making it hard to perceive.
Fig.~\ref{fig:latency} also plots the latency values for the baseline
``SC'' approach which achieves convergence by synchronizing each
operation, i.e., executing under strong consistency (SC). The SC write
latency increases super-linearly with the median value of 1.76s for 60
concurrent editors. The maximum latency measured for an SC write is
9.02s, which is considerably more than the inter-edit latency of
0.25s. The experiment confirms the theory that convergence does not
come at the cost of latency in
\quark.

@@

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with the progression of
time, $B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't preempt it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators in the multiples of 3 until 30. While
increased staleness is inevitable in our approach due to synchronized
merges, we believe the increase can be contained by choosing the order
of merges to avoid the ``starvation'' of some branches. Note that
relaxing the linearizability constraint on merges would completely
eliminate the staleness overhead, but the resultant Git-like system
fails to converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits anamolous version history graphs where two branches can
  have the same set of commits and yet differ in their final version.
  Supplementary material describes two such cases we observed on
  Github.}.

### 2021-11-20 introduction.tex
Edits

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions.  Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to not synchronize the execution of RDT
operations. Orchestrating a well-formed convergent execution without
impacting latency and availability may seem impossible due to the CAP
theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
does rule out linearizability, it does not prohibit enforcing weaker
consistency constraints on distributed executions. Our approach
exploits this observation by (\rom{1}). Identifying a novel set of
constraints on distributed executions that guarantee their
convergence, and (\rom{2}). Localizing such constraints on
(asynchronously executed) state merges such that per-operation latency
and system-wide availability remain unaffected.
Sec.~\ref{sec:concrete-sem} explains our runtime in detail.

>>>>

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions.  Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to not synchronize the execution of RDT
operations. Orchestrating a well-formed convergent execution without
impacting latency and availability may seem impossible due to the CAP
theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
does rule out linearizability (as perceived by the user), it does not
prohibit enforcing weaker consistency constraints on distributed
executions. Our approach exploits this observation by (\rom{1}).
Identifying a novel set of constraints on distributed executions that
guarantee their convergence, and (\rom{2}). Localizing such
constraints on (asynchronously executed) state merges such that
per-operation latency and system-wide availability remain unaffected.
Sec.~\ref{sec:concrete-sem} explains our runtime in detail.

### 2021-11-20 related.tex
Edits

The idea of state-based replicated data types with three-way merge
functions was first proposed in~\cite{mrdt}, where authors demonstrate
an approach for deriving merge functions from first principles for
various data types. However, the resultant \emph{mergeable replicated
data types} (MRDTs) were never proved to be convergent. As we
demonstrate for the \C{Set} data type in Sec.~\ref{sec:motivation},
simple merge functions with intuitive merge semantics can nonetheless
lead to divergent executions. The \quark preempts such executions to
guarantee convergence on MRDTs.

@@

Finally, the implementation of \quark bears resemblance to
Quelea~\cite{pldi15}. Both these systems are implemented as shim
layers on top of off-the-shelf distributed stores, and both offload
low-level concerns to the underlying store. However, both these
systems serve orthogonal purposes. Quelea enforces high-level
invariants on applications built with CRDTs. Operations are assumed to
be commutative and applications are assumed to be convergent. In
contrast, \quark enforces convergence of data types that are otherwise
not guaranteed to converge.

>>>>

The idea of replicated data types with three-way merge functions was
first proposed in~\cite{mrdt}, where authors demonstrate an approach
for deriving merge functions from first principles for various data
types. However, the resultant \emph{mergeable replicated data types}
(MRDTs) were never proved to be convergent. As we demonstrate for
the \C{Set} data type in Sec.~\ref{sec:motivation}, simple merge
functions with intuitive merge semantics can nonetheless lead to
divergent executions. \quark can thus be used with MRDTs to preempt
such executions and guarantee their convergence.

@@

Finally, the implementation of \quark bears resemblance to
Quelea~\cite{pldi15}. Both these systems are implemented as shim
layers on top of off-the-shelf distributed stores, and both offload
low-level concerns to the underlying store. Quelea however enforces
high-level invariants on applications built with CRDTs. Operations are
assumed to be commutative and applications are assumed to be
convergent. In contrast, \quark enforces convergence of data types
that are otherwise not guaranteed to converge.
